VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms

### 2019-03-18 paper/main.tex
finalizing changes based on feature selection

\begin{abstract}
\noindent\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls. We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
\\
\textbf{Availability:} The scripts used to generate results presented in this paper are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\end{abstract}

@@

\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls. % to show how the classifiers tend to outperform the other methods.  
We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
%this was probably too detailed, need a different sentence
%The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  
\\
\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Given a prioritized list of variants, analysts manually inspect those variants and curate a list of variants to ultimately report to the ordering physician.  Unfortunately, manual curation is time consuming and exhausting.  An analyst must inspect each variant and the associated metadata while simultaneously maintaining a mental picture of the patient's phenotype, leading to what we colloquially refer to as ``variant fatigue".  This variant fatigue means that variants at the end of the prioritized list are often regarded with less scrutiny and/or a less accurate patient model than those near the beginning.  Methods that can prioritize these variants accurately can reduce the impact of variant fatigue, reducing the chances that variants are overlooked or mis-identified.  Additionally, if causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.  Finally, accurate prioritization is a step towards the ultimate goal of automatically identifying all variants that cause a patient's primary phenotypes.

One of the issues with previously published ranking methods is that they were primarily tested on simulated datasets with known, single-gene, pathogenic variants injected into real or simulated background genomic datasets.  Additionally, when phenotype terms were used, they tended to use all available phenotype terms paired with the simulated disease with a few noisy terms added or removed.

In this paper, we focus on real patient data from the multi-site collaboration of the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.  Patients accepted into the UDN are believed to have rare, undiagnosed diseases of genetic origin.  Because the UDN is not focused on a single particular disease, the patient population has a diverse range of phenotypes represented.  Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as phenotypes caused by old age).  
%Finally, analysts have the discretion to send back variants with different levels of pathogenicity as specified by the ACMG guidelines \citep{richards2015}.  
Because the UDN is a research collaboration, there is also variability in reported variants that range in pathogenicity from ``variant of uncertain significance" (VUS) through ``pathogenic" as defined by the ACMG guidelines \citep{richards2015}. 
The summation of this real-world variation means that accurately prioritizing variants is challenging due to noise and variation in phenotype inputs and variation in pathogenicity of reported variant outputs.

\section{Approach}
In this paper, we tested the application of classification algorithms for two purposes: 1) predicting whether a variant observed by an analyst would be clinically reported and 2) prioritizing all variants seen by clinical analysts.  In particular, we focused our analyses on real patients with a diverse collection of rare, undiagnosed diseases that were admitted to the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.  We limited our patients to those who received whole genome sequencing and received at least one primary variant (i.e. not secondary or incidental) on their clinical report.  We extracted data directly from the same annotation and filtering tool used by the analysts in order to replicate their data view of each variant in a patient.  Additionally, we incorporated phenotype information into the models using two scoring systems that are based on ranking genes by their association to a set of patient phenotypes.  Finally, each variant was either labeled as ``returned" or ``not returned" depending on whether it was ultimately reported back to the clinical site.

Given the above variant information, we split the data into training and testing sets for measuring the performance of classifiers to predict whether a variant would be clinically reported or not.  We tested four classifiers that are readily available in the {\it sklearn} \citep{pedregosa2011} and {\it imblearn} \citep{lemaitre2017} Python modules.  Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing for both classification analysis and ranking of the variants by their calculated probabilities.  After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant ranking perspective.  All of the scripts to train classifiers, test classifiers, and format results are contained in the VarSight repository.

@@

The first method uses phenotype-to-gene annotations provided by the HPO to calculate a cosine score \citep{kohler2017} between the patient's phenotypes and each gene.  In short, this method builds a vector from the phenotype terms (and ancestor terms in the ontology) for both the patient and each gene.  Then, the cosine between the patient vector and each gene vector is calculated as a representation of similarity.  This method tends to be more conservative because it relies solely on curated annotations from the HPO.

The second method, an internally-developed tool called PyxisMap \citep{wilk2018}, uses the same phenotype-to-gene annotations from the HPO, but adds in automatically text-mined annotations from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.  The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.  We used PyxisMap v1.2, and we ran the standard installation script that downloads all required data sources on December 19, 2018.  Each method generates a single numerical feature that is used in the following analyses.

\subsection{Patient selection}
In our analysis, we focused on variants that were clinically reported as ``primary", meaning the analysts believed the variant to be directly related to the patient's phenotype.  Note that secondary and/or incidental findings are specifically not included in this list.  The analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic adhering to the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.

@@

\subsection{Data cleaning}
For the purposes of classification, all annotations needed to be cleaned and stored as numerical features.  For numerical annotations (e.g. float values like CADD or GERP), we simply copied the annotation over as a single value feature.  Missing annotations were assigned a default value that was outside the expected value range for that feature.  Additionally, these default values were always on the less impactful side of the spectrum (e.g. a default conservation score would err on the side of not being conserved).  The one exception to this rule was for variant allele frequencies where a variant absent from the database was considered to have an allele frequency of 0.0.

For categorical data, we used a two step approach to cleaning the data: bin-count encoding and principal component analysis.  First, we chose to use a bin-count because there are many categories where multiple categorical labels may be present at different quantities.  For example, a single ClinVar variant may have multiple entries where different sites have selected different levels of pathogenicity.  In this situation, we desired to capture not only the categorical label as a feature, but also the number of times that label occurred in the annotations.  Second, we found that the bin-count encoding tended to create many extra features (one per category per annotation type) that were ignored and/or diluted the useful features from the pool of features.  To reduce this dilution, we used principal component analysis \citep{jolliffe2011} to reduce the dimensions of each category and stored at most two features per categorical feature.  %For specifics on variance explained by the top two PCA components per category, please see Supplementary document X (If we ever do this, it will be at the request of a user or reviewers, it doesn't seem super important at this juncture)

@@

We then selected four publicly available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.  The primary reason for selecting these four classifiers is that they are all capable of handling imbalanced data either by default or with a single parameter specifying to balance the data during training.  The two random forest models build collections of decision trees that weight each training input by its class frequency.  Logistic regression calculates the probability of a value belonging to a particular class, again weighting by the class frequency.  In contrast to the other three tested methods, the ensemble classification balances the training input using random under-sampling and then trains an ensemble of AdaBoost learners.

Initially, we also tried the support vector classifier by {\it sklearn}, the multi-layer perceptron by {\it sklearn}, and the random under-sampling AdaBoost classifier by {\it imblearn}.  Each of these was excluded from our results due to, respectively, scaling issues with the training size, failure to handle the data imbalance, and overfitting to the training set.  While we did not achieve good results using these out-of-the-box implementations, it may be possible to use the methods with appropriate changes made to the underlying implementations.

For each of our tested models, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.  For each model and set of hyperparameters, we performed 10-fold cross validation on the training variants and recorded the balanced accuracy scores (i.e. a weighted accuracy score where inputs are weighted by their inverse class frequency).  For each model type, we saved the hyperparameters and model with the best average balanced accuracy score.  These four tuned models were then tested against the unseen set of test proband cases.
\ifpreprint

@@

\section{Results}%mana said results!
\subsection{Classifier Statistics}
\begin{figure*}
\centerline{
\includegraphics[width=0.45\textwidth]{codi_rf_roc.png}
\includegraphics[width=0.45\textwidth]{codi_rf_pr.png}
}
\caption{Receiver operator curves.  
%This figure shows the receiver operator curves for the test data for each single-measure statistic and for each trained classifier.  
This figure shows the receiver operator curves for the test data for each trained classifier.  
On the left, we show the false positive rate against the true positive rate.  On the right, we show the recall against the precision.  Area under the receiver-operator curve (AUROC) is reported beside each method in the legend.  In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly with LogisticRegression performing slightly worse overall.}
\label{fig:roc}
\end{figure*}

For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.  We then calculated the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC) using the unseen test set (see Table \ref{tab:classifier_results}).  Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the four models.

From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.  However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was 0.2109).  This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.

\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}
\caption{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.}
\label{tab:classifier_results}
\end{table}

@@

\begin{table}[!t]
\processtable{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures. \label{tab:classifier_results}}{
\begin{tabular}{c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi

\ifpreprint
\newgeometry{margin=1cm} % modify this if you need even more space
\begin{landscape}
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}
\caption{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  
%Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  
The middle four rows (two Exomiser runs, Phen-Gen, and DeepPVP) represent external tools that ranked the same set of variants as the classifier algorithms.  Phen-Gen was the only external tool that did not rank every variant in the set, so we conservatively assumed unranked variants were at the next best position despite being unranked.
The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values in this table were generated using only the test data that was unseen during training.}
\label{tab:performance}
\end{table}
\end{landscape}
\restoregeometry
\subsection{Ranking Statistics}

@@

\subsection{Ranking Statistics}
\begin{table*}[!t]
\centering
\processtable{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  
%Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  
The middle four rows (two Exomiser runs, Phen-Gen, and DeepPVP) represent external tools that ranked the same set of variants as the classifier algorithms.  Phen-Gen was the only external tool that did not rank every variant in the set, so we conservatively assumed unranked variants were at the next best position despite being unranked.
The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values in this table were generated using only the pre-filtered variants from testing data. \label{tab:performance}}{
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}}{}%{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
\end{table*}
\fi

@@

For comparison, we selected to run Exomiser \citep{smedley2015a}, Phen-Gen \citep{javed2014}, and DeepPVP \citep{boudellioua2019}.  For each tool, we input the exact same set of phenotype terms used by the classifiers we tested.  Additionally, we used the pre-filtered variant set from Codicem as input to each ranking algorithm.  As a result, all external tools and our trained classifiers are ranking on identical phenotype and variant information.

>>>>

\begin{abstract}
\noindent\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls. We showed that these classifiers outperformed the other methods with the best classifiers ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
\\
\textbf{Availability:} The scripts used to generate results presented in this paper are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\end{abstract}

@@

\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls. % to show how the classifiers tend to outperform the other methods.  
We showed that these classifiers outperformed the other methods with the best classifier ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
%this was probably too detailed, need a different sentence
%The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  
\\
\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Given a prioritized list of variants, analysts manually inspect those variants and curate a list of variants to ultimately report to the ordering physician.  Unfortunately, manual curation is a time consuming process where analysts must inspect each variant while maintaining a mental picture of the patient's phenotype.  \citealp{bick2017} report an average of 600 variants per case analyzed by two people (one analyst and one director) over three hours ($\approx$100 variants per man-hour).  If causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.  Finally, accurate prioritization is a step towards the ultimate goal of automatic identification of rare variants that may cause a patient's primary phenotypes.

One of the issues with previously published ranking methods is that they were primarily tested on simulated datasets with known, single-gene, pathogenic variants injected into real or simulated background genomic datasets.  Additionally, when phenotype terms were used, they tended to use all available phenotype terms paired with the simulated disease with a few (typically 2-3) noisy terms added or removed.  Unfortunately, rare disease patient data is often far noisier due to a wide variety of reasons.

In this paper, we focus on real patient data from the multi-site collaboration of the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.  Patients accepted into the UDN are believed to have rare, undiagnosed diseases of genetic origin.  Because the UDN is not focused on a single particular disease, the patient population has a diverse range of phenotypes represented.  Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as phenotypes caused by old age).  Because the UDN is a research collaboration, there is also variability in reported variants that range in pathogenicity from ``variant of uncertain significance" (VUS) through ``pathogenic" as defined by the ACMG guidelines \citep{richards2015}. The summation of this real-world variation means that accurately prioritizing variants is challenging due to noise and variation in phenotype inputs and variation in pathogenicity of reported variant outputs.

\section{Approach}
We tested the application of classification algorithms for identifying clinically reported variants in real world patient in two ways: 1) predicting whether a variant observed by an analyst would be clinically reported and 2) prioritizing all variants seen by the clinical analysts.  In particular, we focused our analyses on real patients with a diverse collection of rare, undiagnosed diseases that were admitted to the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.  We limited our patients to those who received whole genome sequencing and received at least one primary variant (i.e. not secondary or incidental) on their clinical report.  We extracted data directly from the same annotation and filtering tool used by the analysts in order to replicate their data view of each variant in a patient.  Additionally, we incorporated phenotype information into the models using two scoring systems that are based on ranking genes by their association to a set of patient phenotypes.  Finally, each variant was either labeled as ``returned" or ``not returned" depending on whether it was ultimately reported back to the clinical site.

Given the above variant information, we split the data into training and testing sets for measuring the performance of classifiers to predict whether a variant would be clinically reported or not.  We tested four classifiers that are readily available in the {\it sklearn} \citep{pedregosa2011} and {\it imblearn} \citep{lemaitre2017} Python modules.  Of note, our focus was not on picking the ``best'' classifier, but rather on analyzing their overall ability of classification algorithms to handle the noise of real-world patient cases from the UDN.

Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing us to measure their performance as a classifier and as a ranking system.  After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant ranking perspective.  Additionally, we tested four publicly available variant ranking algorithms and two single-value ranking methods for comparison.  All of the scripts to train classifiers, test classifiers, and format results are contained in the VarSight repository.

@@

The first method uses phenotype-to-gene annotations provided by the HPO to calculate a cosine score \citep{kohler2017} between the patient's phenotypes and each gene.  In short, this method builds a vector from the phenotype terms (and ancestor terms in the ontology) for both the patient and each gene.  Then, the cosine of the angle between the patient vector and each gene vector is calculated as a representation of similarity.  This method tends to be more conservative because it relies solely on curated annotations from the HPO.

The second method, an internally-developed tool called PyxisMap \citep{wilk2018}, uses the same phenotype-to-gene annotations from the HPO, but adds in automatically text-mined annotations from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.  The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.  
%We used PyxisMap v1.2, and we ran the standard installation script that downloads all required data sources on December 19, 2018. 
Each method generates a single numerical feature that is used in the following analyses.

\subsection{Patient selection}
In the clinical analysis, each patient was fully analyzed by one director and one analyst.  After the initial analysis, the full team of directors and analysts review flagged variants and determine their reported pathogenicity.  In our analysis, we focused on variants that were clinically reported as ``primary", meaning the team of analysts believed the variant to be directly related to the patient's phenotype.  Note that secondary and/or incidental findings are specifically not included in this list.  The team of analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic adhering to the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.

@@

\subsection{Feature Selection}
\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}
\caption{Feature selection.  This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.  After training, the two random forest models report the importance of each feature in the model (total is 1.00 per classifier).  We average the two importance values, and order them from most to least important.  Feature labels with an `\_' represent a single category of a multi-category feature (i.e. ``HGMD assessment type\_DM'' means the ``DM'' bin-count feature from the ``HGMD assessment type'' annotation in Codicem).}
\label{tab:features}
\end{table}

\begin{table}[!t]
\processtable{Feature selection.  This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.  After training, the two random forest models report the importance of each feature in the model (total is 1.00 per classifier).  We average the two importance values, and order them from most to least important.  Feature labels with an `\_' represent a single category of a multi-category feature (i.e. ``HGMD assessment type\_DM'' means the ``DM'' bin-count feature from the ``HGMD assessment type'' annotation in Codicem). \label{tab:features}}{
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi

For the purposes of classification, all annotations needed to be cleaned and stored as numerical features.  For single-value numerical annotations (e.g. float values like CADD or GERP), we simply copied the annotation over as a single value feature.  Missing annotations were assigned a default value that was outside the expected value range for that feature.  Additionally, these default values were always on the less impactful side of the spectrum (e.g. a default conservation score would err on the side of not being conserved).  The one exception to this rule was for variant allele frequencies where a variant absent from the database was considered to have an allele frequency of 0.0.  For multi-value numerical annotations, we reduced the values (using minimum or maximum) to a single value that was used as the feature.

%this was the old way before feature selection
%For categorical data, we used a two step approach to cleaning the data: bin-count encoding and principal component analysis.  First, we chose to use a bin-count because there are many categories where multiple categorical labels may be present at different quantities.  For example, a single ClinVar variant may have multiple entries where different sites have selected different levels of pathogenicity.  In this situation, we desired to capture not only the categorical label as a feature, but also the number of times that label occurred in the annotations.  Second, we found that the bin-count encoding tended to create many extra features (one per category per annotation type) that were ignored and/or diluted the useful features from the pool of features.  To reduce this dilution, we used principal component analysis \citep{jolliffe2011} to reduce the dimensions of each category and stored at most two features per categorical feature.  The specific features used and the feature types are detailed in the Supplementary Documents.

For categorical data, we relied on bin-count encoding to store the features.  We chose to bin-count because there are many annotations where multiple categorical labels may be present at different quantities.  For example, a single ClinVar variant may have multiple entries where different sites have selected different levels of pathogenicity.  In this situation, we desired to capture not only the categorical label as a feature, but also the number of times that label occurred in the annotations.

After converting all annotations to numerical features or multiple bin-count encoded features, we had a total of 95 features per variant.  We then pruned down to only the top 20 features using univariate feature selection (specifically the \texttt{SelectKBest} method of \texttt{sklearn}).  This method evaluates how well an individual feature performs as a classifier and keeps only the top 20 features for the full classifiers.  Table \ref{tab:features} shows the list of retained features ordered by feature importance after training.  Feature importance was derived from the random forest models which automatically report how important each feature was to classification.  The entire set of annotations along with descriptions of how each was processed prior to feature selection are detailed in the Supplementary Documents.

@@

We then selected four publicly available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.  These classifiers were chosen for three main reasons: 1) their ability to handle imbalanced data (i.e. far more unreported variants than reported variants), 2) their ability to scale to the size of the training and testing datasets, and 3) they are freely available implementations that can be tuned and tested with relative ease.  The two random forest models build collections of decision trees that weight each training input by its class frequency.  Logistic regression calculates the probability of a value belonging to a particular class, again weighting by the class frequency.  In contrast to the other three tested methods, the ensemble classification balances the training input using random under-sampling and then trains an ensemble of AdaBoost learners.  For more details on each classifier, please refer to the {\it sklearn} and {\it imblearn} documentations.

Initially, we also tested the support vector classifier by {\it sklearn}, the multi-layer perceptron by {\it sklearn}, and the random under-sampling AdaBoost classifier by {\it imblearn}.  Each of these was excluded from our results due to, respectively, scaling issues with the training size, failure to handle the data imbalance, and overfitting to the training set.  While we did not achieve positive results using these three implementations, it may be possible to use the methods through another implementation.

For each of our tested models, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.  For each model and set of hyperparameters, we performed stratified 10-fold cross validation on the training variants and recorded the balanced accuracy (i.e. weighted accuracy based on inverse class frequency) and the F1 scores (i.e. harmonic mean between precision and recall).  For each model type, we saved the hyperparameters and model with the best average F1 score.  These four tuned models were then trained on the full training set and tested against the unseen set of test proband cases.  The set of hyperparameters tested along with the highest performance setting for each hyperparameter can be found in the Supplementary Documents.

@@

\section{Results}%mana said results!
\subsection{Classifier Statistics}
\begin{figure*}
\centerline{
\includegraphics[width=0.45\textwidth]{codi_rf_roc.png}
\includegraphics[width=0.45\textwidth]{codi_rf_pr.png}
}
\caption{Receiver operator and precision-recall curves.  
These figures show the performance of the four classifiers on the testing set after hyperparameter tuning and fitting to the training set.  On the left, we show the receiver operator curve (false positive rate against the true positive rate).  On the right, we show the precision recall curve.  Area under the curve (AUROC and AUPRC) is reported beside each method in the legend.}
\label{fig:roc}
\end{figure*}

Each model was tuned using 10-fold cross validation and the resulting average and standard deviation of balanced accuracy is reported in Table \ref{tab:classifier_results}.  After fitting the tuned model to the full training set, we evaluated the models on the testing set by calculating the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC).  Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the results from the testing set on all four models.

From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.  However, these classifier all have relatively poor performance from a precision-recall perspective (best AUPRC was 0.2109).  This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.

\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}
\caption{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC). The CV10 Acc. was gathered during hyperparameter tuning by calculating the average and standard deviation of the 10-fold cross validation.  AUROC and AUPRC was evaluated on the testing set after hyperparameter tuning and fitting to the full training set.}
\label{tab:classifier_results}
\end{table}

@@

\begin{table}[!t]
\processtable{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC). The CV10 Acc. was gathered during hyperparameter tuning by calculating the average and standard deviation of the 10-fold cross validation.  AUROC and AUPRC was evaluated on the testing set after hyperparameter tuning and fitting to the full training set. \label{tab:classifier_results}}{
\begin{tabular}{c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi

\ifpreprint
\newgeometry{margin=1cm} % modify this if you need even more space
\begin{landscape}
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}
\caption{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested. The middle four rows (two Exomiser runs, Phen-Gen, and DeepPVP) represent external tools that ranked the same set of variants as the classifier algorithms.  Phen-Gen was the only external tool that did not rank every variant in the set, so we conservatively assumed unranked variants were at the next best position despite being unranked.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (variant of uncertain significance (VUS), likely pathogenic (LP), or pathogenic (Path.)) derived from the ACMG guidelines.  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values in this table were generated using only the Codicem-filtered variants from testing set.}
\label{tab:performance}
\end{table}
\end{landscape}
\restoregeometry
\subsection{Ranking Statistics}

@@

\subsection{Ranking Statistics}
\begin{table*}[!t]
\centering
\processtable{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested. The middle four rows (two Exomiser runs, Phen-Gen, and DeepPVP) represent external tools that ranked the same set of variants as the classifier algorithms.  Phen-Gen was the only external tool that did not rank every variant in the set, so we conservatively assumed unranked variants were at the next best position despite being unranked.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (variant of uncertain significance (VUS), likely pathogenic (LP), or pathogenic (Path.)) derived from the ACMG guidelines.  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values in this table were generated using only the Codicem-filtered variants from testing set. \label{tab:performance}}{
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}}{}%{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
\end{table*}
\fi

@@

For comparison, we selected to run Exomiser \citep{smedley2015a}, Phen-Gen \citep{javed2014}, and DeepPVP \citep{boudellioua2019}.  For each tool, we input the exact same set of phenotype terms used by the classifiers we tested.  Additionally, we used the same set of pre-filtered variants from Codicem as input to each ranking algorithm.  As a result, all external tools and our trained classifiers are ranking on identical phenotype and variant information.
